---
title: "Applying the Genealogical Soritng Index in pairwise fashion"
author: David Winter    
output: pdf_document
---

## Introduction

In our in-review publication [url to preprint to come], we demonstrate the the
way in which the Genealogical Sorting Index (_gsi_) is typically used in species
delimitation studies can generated misleading results. Specifically, the
 _gsi_ value calculated for a given group depends on the structure of the 
 entire tree, meaning the inclusion or exclusion of un-related groups can lead
 to increased _gsi_ values. In our manuscript we suggest one way to account for
 this property while retaining the useful features of _gsi_, which we call the
 pairwise-gsi or _pwgsi_. 

The rationale for taking a pairwise approach is discussed in the manuscript,
here we supplement that discussion by demonstrated how the _pwgsi_ can be
calculated within R using the existing _genealogicalSorting_ package:

## A typical analysis

The repository includes a phylogeny and `imap` table, mapping the individuals in
the tips of that tree to a given group (here labeled from `A` to `E`):

```{r read}
library(genealogicalSorting)
tr <- read.tree("sample.tr")
assignments <- read.table("samples.imap", stringsAsFactors=FALSE)
head(assignments)
```
If we plot the tree with the tips coloured to match the group-assignments we
find that groups `A` and `B` are mono phonetic with respect to all others, but
groups `C`, `D` and `E` are mixed around:


```{r plot, fig.width=5, fig.height=8,}
tip_to_group <- sapply(tr$tip.label, match, assignments[,1])
#RColorBrewer::brewer.pal(name="Set1", n=5)
grps <- assignments[tip_to_group,2]
grp_cols <- c("A" = "#E41A1C", 
              "B" = "#377EB8", 
              "C" = "#4DAF4A", 
              "D" = "#984EA3", 
              "E" = "#FF7F00")
plot(tr, show.tip.label=FALSE)
tiplabels(grps, bg=grp_cols[grps])
```

Even though the "middle" groups here are not monophyletic, they could still be
species in the process of diverging. The _gsi_ is one of several approaches to
testing this hypothesis. Using the existing library we would calculate values
for each group in this tree... 

```{r gsi}
gsi_vals <- gsi(tr, 0, grps)
gsi_vals
```
... and use a permutation test to obtain p-values under the null hypothesis that
there are no divergent species _in the tree as a whole_:

```{r permute}
set.seed(321)
permuted <- permutationTest(tr, 0, grps, 1000)
data.frame(group=gsi_vals$group,
           gsi=gsi_vals$gsi, 
           pval=permuted$pval)
```

## A pairwise analysis

So, in the above case we found that, in addition to the monophyletic groups,  there
is a significant signal of exclusive ancestry for groups `D` and `C`.  But the
null hypothesis being tested by that p-value relates to the entire tree, and we
already new that groups `A` and `B` where monophyletic. One way to narrow our
focus down to the groups that we are less sure about is calculate the _gsi_
in pairwise-fashion. To do this, we first drop all the tips that are not part
of the groups being compared ...

```{r drop}
keepers <-  c("C", "D")
subtree <- drop.tip(tr, tr$tip.label[!(grps %in% keepers)])
sub_grps <- grps[grps %in% keepers]
plot(subtree, show.tip.label=F)
tiplabels(sub_grps, bg=grp_cols[sub_grps])
```

... then calculate _gsi_ for this pair of groups on this subtree:

```{r pwgsi}
gsi_vals <- gsi(subtree, 0, sub_grps)
permute <- permutationTest(subtree, 0, sub_grps, 1000)
data.frame(group=gsi_vals$group, 
           gsi=gsi_vals$gsi, 
           pval=permute$pval)
```

The signal of exlusive ancestry that seemed to exist for these groups has 
dissipated now that we've removed the well resolved groups from the tree. 

## Making the _pwgsi_ easy to calculate

As you can see from the code above, the process of dropping tips and selecting
only the group-labels associated with a given pairwise test takes some work. We
can this process easier by writing functions to perform the necessary steps. Note, 
we are now calculating the mean of the two _gsi_ values in a pairwise comparison
(the _pwgsi_):


```{r make_fxn}
pw_permute <- function(tree, assignments){
    assignments <- sample(assignments)
    mean(gsi(tree, 0, assignments)$gsi)
}

pwgsi <- function(groups, tree, assignments, nperm=1000){
    if(length(groups) !=2) {
        stop("argument 'groups' must specify exactly two groups for pwgsi")
    }
    subtree <- drop.tip(tree, tree$tip.label[!(assignments %in% groups)])
    sub_grps <- grps[assignments %in% groups]
    pwgsi <- mean(gsi(subtree, 0, sub_grps)$gsi)
    null_distr <- replicate(nperm, pw_permute(tree, assignments))
    list(Grp1 = groups[1], 
         Grp2=groups[2], 
         pwgsi=pwgsi, 
         pval=mean(pwgsi <= null_distr))
}

pwgsi_analysis <- function(tree, assignments, nperm=1000, focal_grps=NULL){
    if(is.null(focal_grps)){
        focal_grps <- unique(assignments)
    }
    to_compare = combn(focal_grps, 2)
    cat("Calculating pwgsi for", dim(to_compare)[2], "pairwise-comparisons...\n")
    res <- apply(to_compare, 2, function(g) pwgsi(g, tree, assignments, nperm))
    do.call(rbind.data.frame, res)    
}
```

The most important of these function is `pwgsi_analysis`, which calculates  
_pwgsi_ for two-group combinations. It takes a complete tree (i.e with no
dropped tips), a vector of group assignments (such as can be produced by
`readAssisgnmentFile` from the `genealogicalSorting` package) and the number of permutations
from which to calculate p-values as input. The optional argument `focal_grps`
specifies the groups from which pairwise comparisons will be made. If
has more than two groups, the functional will calculate _pwgsi_ for 
all unique group-pairs. So, for instance, can calculate _pwgsi_ using all groups
other than the well-resolved `A` and `B`:


```{r, demo}
pwgsi_analysis(tr, grps, 1000, c("C", "D", "E"))
```

If `focal_grps` is not specified, this function calculates all possible 
pairwise comparisons. With a little bit of fussing, we can use the `data.frame`
returned by `pwgsi_analysis` to plot of _pwgsi_ visualising the degree of
exclusive ancestry exhibited been all possible group-pairs in this example: 


```{r matix}
library(ggplot2)
res <- pwgsi_analysis(tr, grps, 100)

res$Grp1 <- factor(res$Grp1, unique(as.character(res$Grp1)))
res$Grp2 <- factor(res$Grp2, rev(unique(as.character(res$Grp2))))
ggplot(res, aes(Grp1, Grp2, fill=pwgsi)) + geom_tile(colour="white")
```

## Using these functions. 

This contents of this file are made freely available under a
[CC0](https://creativecommons.org/publicdomain/zero/1.0/) license, so you are
free to use the specific functions  defined about in your analyses. If you do
so, please cite both our paper [citation to come] and [Cummings 2008](http://10.1111/j.1558-5646.2008.00442.x) for the original description of the _gsi_.
